---
title: "Example 1: RCBD & Negative Block Effects"
bibliography: ../references.bib
css: [../style.css, ../toc.css]
---

```{r,echo=FALSE, include=FALSE, message=FALSE}
library(rmarkdown); library(kableExtra)
library(SASmarkdown)
saspath <- "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
sasopts <- "-nosplash -ls 200"
```

::: callout-note
## GLMM Challenge

How to handle negative variance component estimates for block effects.
:::

## Study Design

-   Randomized complete block design with 5 blocks
-   6 treatments randomly assigned to units within each block
-   Gaussian response variable

Example from @stroup_sas_2018, Ch 2, Section 2.5.

## Data Generation Process

```{r, echo=FALSE}
table1 <- data.frame(
  col1 = c("Block", "", "Unit(block)", "Total"),
  col2 = c( 4, "", 25, 29),
  col3 = c("", "Treatment", "", ""),
  col4 = c("", 5, "", ""),
  col5 = c("Block", "Treatment", "Unit(block)|Trt*", "Total"),
  col6 = c(4, 5, '25-5=20', 29)
)

kable(table1, col.names = NULL, booktabs = TRUE, format = "html") |> 
  add_header_above(rep(c("Source", "df"), 3), align = "l") |> 
  add_header_above(c("Study Design" = 2, "Treatment Design" = 2, "Combined" = 2)) |> 
  row_spec(4, bold = TRUE) |> 
  kable_styling(full_width = TRUE, position = "center")
```

*\*Unit-within-block effects after accounting for treatment effects.*

::: {.callout-caution collapse="false"}
## Important note

@stroup_generalized_2024 define a “sensible model” as one for which there is a one-to-one correspondence between sources of variation and model effects that account for said sources. “Not sensible” models either fail to account for one or more sources, or have two or more model terms that account for the same source and are therefore confounded.
:::

## Statistical Model

-   Linear predictor: $$ \eta_{ij} = \eta + \tau_i + \beta_j$$

-   Distributions:

Blocks effects, $b_j$, are iid and

$$ b_j \sim N(0, \sigma^2_b) $$
       
Observations: 

$$ y_{ij}|\beta_j \sim N(\mu_{ij}, \sigma^2) $$

-   Link function:

$$ \eta_{ij} = \mu_{ij} $$

-   Treatment mean:

$$ E(y_{ij}) = \mu_{i} $$

## GLMM Challenge(s) in Detail

As a standard Gaussian mixed model, no complications are expected. However, in this case, the variance estimating equation yields a negative solution. A common default software solution to is force the variance, $\hat{\sigma^2_b}$, to be zero. The set-to-zero approach effectively pools block-level and unit-level sources of variation, resulting in:

-   a downward biased residual variance estimate;
-   an upward biased test statistics;
-   an inflated type I error rate; and
-   reduced accuracy of confidence interval coverage.

See Stroup and Littell (2002) for a detailed discussion of this issue and for justification of the recommended best practice illustrated in this example.

## Analytical Solutions

### R

```{r, echo=TRUE, message=FALSE, warning=FALSE, include=FALSE}
library(here); library(nlme); library(performance); library(lme4)
library(emmeans)
```

#### Import Data and Examine

```{r}
example_1 <- read.csv(here::here("data", "example_01.csv")) 

example_1$Trt <- as.factor(example_1$Trt)
example_1$Block <- as.factor(example_1$Block)
```

```{r}
table(example_1$Block, example_1$Trt)
hist(example_1$Y)
```

#### Original Analysis

*Note that the following two models give the same results.*

```{r}
example1_m1 <- lme4::lmer(Y ~ Trt + (1|Block),
                   data = example_1,
                   REML = TRUE)
```

```{r, echo=TRUE, eval=FALSE}
example1_m1 <- lme(Y ~ Trt,
                   random = ~1|Block,
                   data = example_1,
                   method = "REML")
```

Examine variance components and test for singularity (i.e. a zero variance component)

```{r}
print(VarCorr(example1_m1), comp = "Variance")
check_singularity(example1_m1)
```

#### Alternative

Fit the model using a compound symmetry variance structure, which will allow negative correlations within groups.

```{r}
example1_m2 <- gls(Y ~ Trt, 
                   data = example_1,
                   method = "REML",
                   corr = corCompSymm(form = ~ 1|Block, fixed = FALSE))
```

```{r}
summary(example1_m2)
check_singularity(example1_m2)
```

```{r}
joint_tests(example1_m1)
joint_tests(example1_m2)
```

### SAS

#### Import data

```{r collapse=TRUE, engine="sashtml", collectcode=TRUE, engine.path=saspath, include=TRUE, engine.opts=sasopts, comment=""}
data Example1Data;
   input block trt Y;
datalines;
1 1 24.739
1 2 22.719
1 3 27.247
1 4 28.45
1 5 30.202
1 6 30.422
2 1 27.921
2 2 29.419
2 3 27.025
2 4 30.467
2 5 23.334
2 6 23.299
3 1 24.064
3 2 22.945
3 3 26.169
3 4 30.414
3 5 28.661
3 6 29.571
4 1 25.102
4 2 28.342
4 3 29.337
4 4 23.714
4 5 33.414
4 6 26.563
5 1 22.352
5 2 30.348
5 3 23.798
5 4 29.197
5 5 30.625
5 6 27.528
;
run;
```

#### Original Analysis

```{r engine="sashtml", engine.path=saspath, include=TRUE, engine.opts=sasopts, comment=""}
proc glimmix data=Example1Data;
   class block trt;
   model y = trt;
   random block;
run;
```

#### Alternative Approach #1

Fit the model using a compound symmetry variance structure, which will allow negative correlations within groups.

```{r engine="sashtml", engine.path=saspath, include=TRUE, engine.opts=sasopts, comment=""}
proc glimmix data=Example1Data;
   class block trt;
   model y = trt;
   random trt / subject=block type=cs residual v vcorr;
run;
```

#### Alternative Approach #2

Fit the model allowing the variance componens to become negative.

SAS (unlike R) has an option for overriding the restriction on negative variance components.

```{r engine="sashtml", engine.path=saspath, include=TRUE, engine.opts=sasopts, comment=""}
proc glimmix data=Example1Data nobound;
   class block trt;
   model y = trt;
   random block;
run;
```

## References:

::: {#refs}
:::
