---
title: "Example 1: RCBD & Negative Block Effects"
subtitle: "Solution in SAS"
css: [../css/sas.css]
---


```{r,echo=FALSE, include=FALSE, message=FALSE}
require(SASmarkdown)
saspath <- "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
sasopts <- "-nosplash -ls 200"
knitr::opts_chunk$set(engine.path=list(sas=saspath, saslog=saspath),
                  engine.opts=list(sas=sasopts, saslog=sasopts), 
                  comment=NA)
```

::: callout-note
## GLMM Challenge

How to handle negative variance component estimates for block effects.
:::


## Import data


```{r collapse=TRUE, engine="sashtml", engine.path=saspath, include=TRUE, engine.opts=sasopts, comment=""}
data Example1Data;
   input block trt Y;
datalines;
1 1 24.739
1 2 22.719
1 3 27.247
1 4 28.45
1 5 30.202
1 6 30.422
2 1 27.921
2 2 29.419
2 3 27.025
2 4 30.467
2 5 23.334
2 6 23.299
3 1 24.064
3 2 22.945
3 3 26.169
3 4 30.414
3 5 28.661
3 6 29.571
4 1 25.102
4 2 28.342
4 3 29.337
4 4 23.714
4 5 33.414
4 6 26.563
5 1 22.352
5 2 30.348
5 3 23.798
5 4 29.197
5 5 30.625
5 6 27.528
;
run;
```


## Original Analysis

```{r collapse=TRUE, engine="sashtml",  engine.path=saspath, include=TRUE, engine.opts=sasopts, comment=""}
proc glimmix data=Example1Data;
   class block trt;
   model y = trt;
   random block;
run;
```



## Alternative Approach 1

Fit the model using a compound symmetry variance structure, which will allow negative correlations within groups.

```{r collapse=TRUE, engine="sashtml", engine.path=saspath, include=TRUE, engine.opts=sasopts, comment=""}
proc glimmix data=Example1Data;
   class block trt;
   model y = trt;
   random trt / subject=block type=cs residual v vcorr;
run;
```



## Alternative Approach 2

Fit the model allowing the variance componens to become negative. SAS has an option for overriding the restriction on negative variance components.

```{r collapse=TRUE, engine="sashtml",  engine.path=saspath, include=TRUE, engine.opts=sasopts, comment=""}
proc glimmix data=Example1Data nobound;
   class block trt;
   model y = trt;
   random block;
run;
```

