---
title: "Example 1: RCBD & Negative Block Effects"
bibliography: ../references.bib
---

```{r,include=FALSE, eval=TRUE}
library(rmarkdown); library(kableExtra)
```

::: callout-note
## GLMM Challenge

How to handle negative variance component estimates for block effects.
:::

## Study Design

-   Randomized complete block design with 5 blocks
-   6 treatments randomly assigned to units within each block
-   Gaussian response variable

Example from @stroup_sas_2018, Ch 2, Section 2.5.

## Data Generation Process

```{r, echo=FALSE}
table1 <- data.frame(
  col1 = c("Block", "", "Unit(block)", "Total"),
  col2 = c( 4, "", 25, 29),
  col3 = c("", "Treatment", "", ""),
  col4 = c("", 5, "", ""),
  col5 = c("Block", "Treatment", "Unit(block)|Ttr*", "Total"),
  col6 = c(4, 5, '25-5=20', 29)
)

kable(table1, col.names = NULL,  format = "html") |> 
  add_header_above(rep(c("Source", "df"), 3), align = "l") |> 
  add_header_above(c("Study Design" = 2, "Treatment Design" = 2, "Combined" = 2)) |> 
  row_spec(4, bold = TRUE) |> 
  kable_styling(full_width = TRUE, position = "center")#|> 
 # footnote("*Unit-within-block effects after accounting for treatment effects.")

```
::: {style="font-size: 75%;"}

*Unit-within-block effects after accounting for treatment effects.

:::


::: {.callout-caution collapse="false"}
## Important note

@stroup_generalized_2024 define a “sensible model” as one for which there is a one-to-one correspondence between sources of variation and model effects that account for said sources. “Not sensible” models either fail to account for one or more sources, or have two or more model terms that account for the same source and are therefore confounded.
:::

## Statistical Model

-   Linear predictor: $$ \eta_{ij} = \eta + \tau_i + \beta_j$$

-   Distributions:

Blocks effects, $b_j$, are iid and

$$ b_j \sim N(0, \sigma^2_b) $$
       
Observations: 

$$ y_{ij}|\beta_j \sim N(\mu_{ij}, \sigma^2) $$

-   Link function:

$$ \eta_{ij} = \mu_{ij} $$

-   Treatment mean:

$$ E(y_{ij}) = \mu_{i} $$

## GLMM Challenge(s) in Detail

As a standard Gaussian mixed model, no complications are expected. However, in this case, the variance estimating equation yields a negative solution. A common default software solution to is force the variance, $\hat{\sigma^2_b}$, to be zero. The set-to-zero approach effectively pools block-level and unit-level sources of variation, resulting in:

-   a downward biased residual variance estimate;
-   an upward biased test statistics;
-   an inflated type I error rate; and
-   reduced accuracy of confidence interval coverage.

## Solutions

1. Fit the model using a compound symmetry variance structure, which will allow negative correlations within groups. This reparametrizes the model as thus:

$$\eta_{ij} = \eta + \tau_i$$


$$ \begin{bmatrix}
y_{1j} \\
y_{2j} \\
y_{3j} \\
y_{4j} \\
y_{5j} \\
y_{6j}
\end{bmatrix} \sim MVN 
\begin{pmatrix} 0, 
\begin{bmatrix} 
    \sigma^2+\sigma_{CS} & \sigma_{CS} & \dots  &\sigma_{CS}\\
       & \ddots &   & \vdots\\
       &  &  \ddots & \sigma_{CS} \\
      & & & \sigma^2+\sigma_{CS} 
    \end{bmatrix}
\end{pmatrix} $$


2. Fit the model allowing the variance components to become negative. This option is available in SAS but not R. 


See Stroup and Littell (2002) for a detailed discussion of this issue and for justification of the recommended best practice illustrated in this example.



## References:

::: {#refs}
:::
